---
title: "Case of study - Data analysis"
author: "Carlos Rojas Cisternas"
date: "14/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## Packages
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(here)
```

## Loading the dataset
```{r}
trips_2021 <- rbind(read.csv(here("Clean data","data_2021_Q1.csv")),
                    read.csv(here("Clean data","data_2021_Q2.csv")),
                    read.csv(here("Clean data","data_2021_Q3.csv")),
                    read.csv(here("Clean data","data_2021_Q4.csv")))
```

## Date format
```{r}
# Started and ended date-time in format yy-mm-dd hh:mm:ss
trips_2021$started_at <- ymd_hms(trips_2021$started_at)
trips_2021$ended_at <- ymd_hms(trips_2021$ended_at)

# Ride lenght in time format hh:mm:ss
trips_2021$ride_length <- hms::as_hms(trips_2021$ended_at - trips_2021$started_at)

# Date without time
trips_2021$date <- date(trips_2021$started_at)

# Year
trips_2021$year <- year(trips_2021$started_at)

# Month
trips_2021$month <- month(trips_2021$started_at,
                          label=TRUE)
levels(trips_2021$month) <- c("Jan","Feb","Mar","Apr","May","Jun",
                              "Jul","Aug","Sep","Oct","Nov","Dec")

# Day of month (1 to 31)
trips_2021$month_day <- day(trips_2021$started_at)


# Day of week (Monday to Sunday)
trips_2021$week_day <- wday(trips_2021$started_at,
                            label=TRUE,
                            week_start=getOption("lubridate.week.start", 1))
levels(trips_2021$week_day) <- c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")

# Hour of day (0 to 23)
trips_2021$hour <- hour(trips_2021$started_at)
```

# Data analysis

## General overview
```{r message=FALSE, warning=FALSE}
trips_time_serie <- trips_2021 %>% 
  select(date, member_casual) %>% 
  group_by(date, member_casual) %>%
  summarize(n_trips=n()) %>% 
  pivot_wider(names_from=member_casual ,values_from = n_trips) %>% 
  mutate(all=member+casual) %>% 
  pivot_longer(!date, names_to="type_customer", values_to="n_trips")

write.csv(trips_time_serie,
          here("Tables","n_trips_time_serie.csv"),
          row.names = FALSE)

ggplot(data = trips_time_serie %>%
         filter(type_customer %in% c("casual","member")) %>% 
         mutate(n_trips=n_trips/1000))+
  geom_line(mapping = aes(x=date,y=n_trips, color=type_customer))+
  scale_color_discrete(name="Customer",
                      labels=c("Casual","Member"))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By type of customer",
       caption="Source: Divvy system data",
       x="Date",
       y="N° trips (in thousands)")+
  theme(legend.position = "bottom")

ggsave(here("Charts","Trips by date and type of customer.jpg"))
```
there is weekly variations between casual and member riders.
there is seasonal variations around the year.

## 1. Analysis of the number of trips

### 1.1 Number of trips by type of customer and day of the week
```{r message=FALSE}
n_customer_dayweek <- trips_2021 %>% 
  select(member_casual, week_day) %>% 
  group_by(member_casual, week_day) %>% 
  summarize(n_trips = n())

write.csv(n_customer_dayweek,
          here("Tables","n_trips_by_customer_dayweek.csv"),
          row.names = FALSE)

ggplot(data = n_customer_dayweek %>%
         mutate(n_trips = n_trips/1000))+
  geom_col(mapping = aes(x=week_day, y=n_trips, fill=member_casual),
           position="dodge")+
  scale_fill_discrete(name="Customer",
                      labels=c("Casual","Member"))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By day of week and type of customer",
       caption="Source: Divvy system data",
       x="Day of week",
       y="N° trips (in thousands)")+
  ylim(0,600)

ggsave(here("Charts","Trips by customer and day of week.jpg"))
```

There are more bike trips on weekends for casual customers, but members ride more between tuesday and thursday.

### 1.2 Number of trips by type of customer and month
```{r message=FALSE}
n_customer_month <- trips_2021 %>% 
  select(member_casual, month) %>%
  group_by(member_casual, month) %>% 
  summarize(n_trips = n())

write.csv(n_customer_month,
          here("Tables","n_trips_by_customer_month.csv"),
          row.names = FALSE)

ggplot(data = n_customer_month %>%
         mutate(n_trips = n_trips/1000))+
  geom_col(mapping = aes(x=month, y=n_trips, fill=member_casual))+
  scale_fill_discrete(name="Customer",
                      labels=c("Casual","Member"))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By month and type of customer",
       caption="Source: Divvy system data 2021",
       x="Month",
       y="N° trips (in thousands)")+
  facet_wrap(~member_casual, nrow=1, ncol=2)+
  theme(legend.position = "bottom")

ggsave(here("Charts","Trips by customer and month.jpg"))
```

Casual customers rides more between june, july and august. These months corresponds to begin of summer in Chicago. Almost the same conclusion for member, they used to ride more between july and september, the summmer months.

### 1.3 Number of trips by type of customer and hour of day

```{r message=FALSE}
n_customer_hour <- trips_2021 %>% 
  select(member_casual, hour) %>%
  group_by(member_casual, hour) %>% 
  summarize(n_trips = n())

write.csv(n_customer_hour,
          here("Tables","n_trips_by_customer_hour.csv"),
          row.names = FALSE)

ggplot(data = n_customer_hour %>%
         mutate(n_trips = n_trips/1000))+
  geom_line(mapping = aes(x=hour, y=n_trips, colour=member_casual),
            size=1)+
  scale_colour_discrete(name="Customer",labels=c("Casual","Member"))+
  scale_x_continuous(breaks = c(0,3,6,9,12,15,18,21))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By hour and type of customer",
       caption="Source: Divvy system data",
       x="Hour of day",
       y="N° trips (in thousands)")

ggsave(here("Charts","Trips by customer and hour of day.jpg"))
```

Both types of customers rides more in the afternoon, between 12 and 18 hrs. However, we can observe that members have two local peaks: The first is around 8 hrs and the second around 17 hrs. On the another hand, casual customers have only 1 local peak at 17 hrs. They doesn't have a peak at 8 hrs. This is probably because members use bikes to get to work in the morning.


### 1.4 Number of trips by customer, hour and day
To figure out more about the last conclusion, I want to see the same chart but by day of the week.

```{r message=FALSE}
n_customer_dayweek_hour <- trips_2021 %>% 
  select(member_casual, week_day, hour) %>%
  group_by(member_casual, week_day, hour) %>% 
  summarize(n_trips = n())

write.csv(n_customer_dayweek_hour,
          here("Tables","n_trips_by_customer_dayweek_hour.csv"),
          row.names = FALSE)

ggplot(data = n_customer_dayweek_hour %>%
         mutate(n_trips = n_trips/1000))+
  geom_line(mapping = aes(x=hour, y=n_trips, colour=member_casual),
            size=1)+
  scale_colour_discrete(name="Customer",labels=c("Casual","Member"))+
  scale_x_continuous(breaks = c(0,3,6,9,12,15,18,21))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By hour, type of customer and day of week",
       caption="Source: Divvy system data",
       x="Hour of day",
       y="N° trips (in thousands)")+
  facet_wrap(~week_day, nrow=4, ncol=3)

ggsave(here("Charts","Trips by customer, hour and day.jpg"))
```

On weekdays, the members have two local peaks of the number of trips, but on weekends both types of customers have the same distribution of the number of trips around the day.

### 1.5 Number of trips by customer, hour and month
```{r message=FALSE}
n_customer_month_hour <- trips_2021 %>% 
  select(member_casual, month, hour) %>%
  group_by(member_casual, month, hour) %>% 
  summarize(n_trips = n())

write.csv(n_customer_month_hour,
          here("Tables","n_trips_by_customer_month_hour.csv"),
          row.names = FALSE)

ggplot(data = n_customer_month_hour %>%
         mutate(n_trips = n_trips/1000))+
  geom_line(mapping = aes(x=hour, y=n_trips, colour=member_casual),
            size=1)+
  scale_colour_discrete(name="Customer",labels=c("Casual","Member"))+
  scale_x_continuous(breaks = c(0,3,6,9,12,15,18,21))+
  labs(title="Chicago bike trips in 2021",
       subtitle="By hour and type of customer",
       caption="Source: Divvy system data",
       x="Hour of day",
       y="N° trips (in thousands)")+
  facet_wrap(~month)

ggsave(here("Charts","Trips by customer, hour and month.jpg"))
```
From this chart, we can extract an important conclusion: mainly on summer months, casual customers rides more than members approximately between 10 and 17 hrs. 

## 2. Analysis of customer's Ride length
### 2.1 Ride lengh by type of customer
```{r}
ride_length_customer <- trips_2021 %>% 
  select(member_casual, ride_length) %>% 
  group_by(member_casual) %>% 
  summarize(average_ride_length = hms::as_hms(round(mean(ride_length),0)))

print(ride_length_customer)
```

On average, causal customers ride more time than members.

### 2.2 Ride length by type of customer and day of week
```{r message=FALSE}
ride_length_customer_dayweek <- trips_2021 %>% 
  select(member_casual, week_day, ride_length) %>% 
  group_by(member_casual, week_day) %>% 
  summarize(average_ride_length = hms::as_hms(round(mean(ride_length),0))) %>%
  pivot_wider(names_from = member_casual, values_from = average_ride_length)

write.csv(ride_length_customer_dayweek,
          here("Tables","ride_length_by_customer_dayweek.csv"),
          row.names = FALSE)

print(ride_length_customer_dayweek)
```

On average, causal customers ride more time than members in all days of week.
Casual riders take longer bike rides on weekends.

### 2.3 Ride lengh by type of customer and hour of day
```{r message=FALSE}
ride_length_customer_hour <- trips_2021 %>% 
  select(member_casual, hour, ride_length) %>%
  group_by(member_casual, hour) %>% 
  summarize(average_ride_length = hms::as_hms(round(mean(ride_length),0))) %>% 
  pivot_wider(names_from = member_casual, values_from = average_ride_length)

write.csv(ride_length_customer_hour,
          here("Tables","ride_length_by_customer_hour.csv"),
          row.names = FALSE)

print(ride_length_customer_hour)
```

On average, causal customers ride more time than members in all hours of day.
Casual riders take longer bike rides between 10am and 3pm

### 2.4 Ride length by type of customer and month
```{r message=FALSE}
ride_length_customer_month <- trips_2021 %>% 
  select(member_casual, month, ride_length) %>%
  group_by(member_casual, month) %>% 
  summarize(average_ride_length = hms::as_hms(round(mean(ride_length),0))) %>% 
  pivot_wider(names_from = member_casual, values_from = average_ride_length)

write.csv(ride_length_customer_month,
          here("Tables","ride_length_by_customer_month.csv"),
          row.names = FALSE)

print(ride_length_customer_month)
```

On average, causal customers ride more time than members in all months of year.
Casual riders take longer bike rides between march and may. These months corresponds to begin of spring in Chicago.

## 3. Analysis of the most popular route trips
### 3.1 Most popular route trips in general
```{r}
popular_routes <- trips_2021 %>% 
  select(start_station_name, end_station_name) %>%
  mutate_all(na_if,"") %>% 
  drop_na() %>% 
  unite(col = route, start_station_name, end_station_name, sep=" to ") %>%
  group_by(route) %>% 
  summarize(n_trips = n()) %>%
  arrange(desc(n_trips))

popular_routes <- head(popular_routes, n=10)
```

The most popular bike routes are round trips in:
- Streeter Dr & Grand Ave
- Michigan Ave & Oak St
- Millennium Park


### By customer
```{r message=FALSE}
customer_popular_routes <- trips_2021 %>% 
  select(start_station_name, end_station_name, member_casual) %>%
  mutate_all(na_if,"") %>% 
  drop_na() %>% 
  unite(col = route, start_station_name, end_station_name, sep=" to ") %>%
  group_by(member_casual, route) %>% 
  summarize(n_trips = n()) %>%
  arrange(desc(member_casual), desc(n_trips))

customer_popular_routes <- rbind(head(filter(customer_popular_routes,
                                             member_casual=="member"),
                                      n=10),
                                 head(filter(customer_popular_routes,
                                             member_casual=="casual"),
                                      n=10))

write.csv(customer_popular_routes,
          here("Tables","popular_routes_by_customer.csv"),
          row.names = FALSE)

```
Members most popular bike routes are:
- from Ellis Ave & 60th St to Ellis Ave & 55th St
- from Ellis Ave & 55th St to Ellis Ave & 60th St
- from Ellis Ave & 60th St to University Ave & 57th St

Casual riders most popular routes are round trips in:
- Streeter Dr & Grand Ave
- Millennium Park
- Michigan Ave & Oak St