---
title: "Case study - Data cleaning"
author: "Carlos Rojas Cisternas"
date: "14/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages
```{r loading packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

# Data cleaning
## Loading data
```{r}
trips_2021_jan <- read.csv(here("Raw data","202101-divvy-tripdata.csv"))
trips_2021_feb <- read.csv(here("Raw data","202102-divvy-tripdata.csv"))
trips_2021_mar <- read.csv(here("Raw data","202103-divvy-tripdata.csv"))
trips_2021_apr <- read.csv(here("Raw data","202104-divvy-tripdata.csv"))
trips_2021_may <- read.csv(here("Raw data","202105-divvy-tripdata.csv"))
trips_2021_jun <- read.csv(here("Raw data","202106-divvy-tripdata.csv"))
trips_2021_jul <- read.csv(here("Raw data","202107-divvy-tripdata.csv"))
trips_2021_aug <- read.csv(here("Raw data","202108-divvy-tripdata.csv"))
trips_2021_sep <- read.csv(here("Raw data","202109-divvy-tripdata.csv"))
trips_2021_oct <- read.csv(here("Raw data","202110-divvy-tripdata.csv"))
trips_2021_nov <- read.csv(here("Raw data","202111-divvy-tripdata.csv"))
trips_2021_dec <- read.csv(here("Raw data","202112-divvy-tripdata.csv"))
```

## Checking if there is duplicates rows
```{r message=FALSE}
datasets <- list(trips_2021_jan,trips_2021_feb,trips_2021_mar,trips_2021_apr,
                 trips_2021_may,trips_2021_jun,trips_2021_jul,trips_2021_aug,
                 trips_2021_sep,trips_2021_oct,trips_2021_nov,trips_2021_dec)

for (k in datasets){
  test_duplicates <- get_dupes(k)

  if(is_empty(test_duplicates$dupe_count)){
    print("There is no duplicate rows")
    remove(test_duplicates)
  } else {
    View(test_duplicates)
    break
  }
}
remove(datasets)
```

## Checking if all the column names are the same
```{r}
compare_df_cols_same(trips_2021_jan,trips_2021_feb,trips_2021_mar,trips_2021_apr,
                     trips_2021_may,trips_2021_jun,trips_2021_jul,trips_2021_aug,
                     trips_2021_sep,trips_2021_oct,trips_2021_nov,trips_2021_dec)
```
## Combining all datasets in one
```{r}
trips_2021 <- rbind(trips_2021_jan,trips_2021_feb,trips_2021_mar,trips_2021_apr,
                    trips_2021_may,trips_2021_jun,trips_2021_jul,trips_2021_aug,
                    trips_2021_sep,trips_2021_oct,trips_2021_nov,trips_2021_dec)

remove(trips_2021_jan,trips_2021_feb,trips_2021_mar,trips_2021_apr,
       trips_2021_may,trips_2021_jun,trips_2021_jul,trips_2021_aug,
       trips_2021_sep,trips_2021_oct,trips_2021_nov,trips_2021_dec)
```

## Cleaning column names and removing possible empty rows
```{r}
trips_2021 <- trips_2021 %>%
  clean_names() %>%
  remove_empty(which = c("rows","cols"))
```


## Variable formats
```{r}
trips_2021$started_at <- ymd_hms(trips_2021$started_at)
trips_2021$ended_at <- ymd_hms(trips_2021$ended_at)
```

## Adding new variables
```{r}
# Ride lenght in time format hh:mm:ss
trips_2021$ride_length <- hms::as_hms(trips_2021$ended_at - trips_2021$started_at)

# Date without time
trips_2021$year <- date(trips_2021$started_at)

# Year
trips_2021$year <- year(trips_2021$started_at)

# Month
trips_2021$month <- month(trips_2021$started_at,
                          label=TRUE)
levels(trips_2021$month) <- c("Jan","Feb","Mar","Apr","May","Jun",
                              "Jul","Aug","Sep","Oct","Nov","Dec")

# Day of month (1 to 31)
trips_2021$month_day <- day(trips_2021$started_at)

# Day of week (Monday to Sunday)
trips_2021$week_day <- wday(trips_2021$started_at,
                            label=TRUE,
                            week_start=getOption("lubridate.week.start", 1))
levels(trips_2021$week_day) <- c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")

# Hour of day (0 to 23)
trips_2021$hour <- hour(trips_2021$started_at)
```

## Summarize of ride length
```{r}
stats_ride_length <- trips_2021 %>%
  select(ride_length) %>% 
  summarise(average = hms::as_hms(round(mean(ride_length),0)),
            median = hms::as_hms(median(ride_length)),
            sd = hms::as_hms(round(sd(ride_length),0)),
            min = hms::as_hms(min(ride_length)),
            max = hms::as_hms(max(ride_length)))
```

There are some problems:
- negative ride length (this doesn't have sense)
- the maximum value is extremly high

Then, I will remove any trips that were below 0 seconds in length. I will also remove any trips that were below 60 seconds because, according to website of Divvy, they are potentially false starts or users trying to re-dock a bike to ensure it was secure.

In the other hand, I will remove any trips that were above 24 hours in length. Although it's possible to use a bike for more time with an additional charge in the tariff, accordingly to the website "If you don't return a bike within a 24-hour period, you may be charged a lost or stolen bike fee of $250 (plus tax)" [link here](https://help.divvybikes.com/hc/en-us/articles/360033484791-What-if-I-keep-a-bike-out-too-long-).

```{r}
trips_2021_cleaned <- trips_2021 %>% 
  filter(ride_length > hms("00:01:00"),
         ride_length <= hms("23:59:59"))
```

Try again

```{r}
stats_ride_length <- trips_2021_cleaned %>%
  select(ride_length) %>% 
  summarise(average = hms::as_hms(round(mean(ride_length),0)),
            median = hms::as_hms(median(ride_length)),
            sd = hms::as_hms(round(sd(ride_length),0)),
            min = hms::as_hms(min(ride_length)),
            max = hms::as_hms(max(ride_length)))
```

Then, we save our data cleaned by quarters (less space)

```{r}
write.csv(trips_2021_cleaned %>% 
            filter(month %in% c("Jan","Feb","Mar")),
          here("Clean data","data_2021_Q1.csv"), 
          row.names=FALSE)

write.csv(trips_2021_cleaned %>% 
            filter(month %in% c("Apr","May","Jun")),
          here("Clean data","data_2021_Q2.csv"), 
          row.names=FALSE)

write.csv(trips_2021_cleaned %>% 
            filter(month %in% c("Jul","Aug","Sep")),
          here("Clean data","data_2021_Q3.csv"), 
          row.names=FALSE)

write.csv(trips_2021_cleaned %>% 
            filter(month %in% c("Oct","Nov","Dec")),
          here("Clean data","data_2021_Q4.csv"), 
          row.names=FALSE)
```